\documentclass[12pt,a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{pgfgantt}
\usepackage{cite}
\usepackage{fancyhdr}
\usepackage{microtype}
\usepackage{caption}

% ── Hyperref setup ────────────────────────────────────────────────────────────
\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=blue!70!black,
    urlcolor=blue!70!black
}

% ── Header / Footer ───────────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\rhead{\small CS Research Proposal --- CoSc3101}
\lhead{\small Hawassa University Institute of Technology}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ── Section formatting ────────────────────────────────────────────────────────
\titleformat{\section}{\large\bfseries\color{blue!80!black}}{}{0em}{\thesection\quad}
\titleformat{\subsection}{\normalsize\bfseries}{}{0em}{\thesubsection\quad}
\titlespacing*{\section}{0pt}{14pt}{6pt}
\titlespacing*{\subsection}{0pt}{10pt}{4pt}

\setlength{\parskip}{4pt}
\setlength{\parindent}{0pt}

% ══════════════════════════════════════════════════════════════════════════════
\begin{document}

% ── Title Page ────────────────────────────────────────────────────────────────
\begin{titlepage}
    \centering
    \vspace*{1cm}
    {\Large \textbf{Hawassa University Institute of Technology}\\[4pt]
    Faculty of Informatics --- Department of Computer Science}\\[0.5cm]
    \rule{\linewidth}{0.5pt}\\[0.5cm]
    {\LARGE \textbf{Detecting and Mitigating Adversarial Attacks\\
    on Deep Learning-Based Intrusion Detection Systems\\
    Using Ensemble Adversarial Training}}\\[0.5cm]
    \rule{\linewidth}{0.5pt}\\[0.8cm]
    {\large \textbf{Research Methods in Computer Science (CoSc3101)}\\[4pt]
    CS Research Proposal --- Individual Project}\\[1cm]
    \begin{tabular}{ll}
        \textbf{Student Name:}   & Dagim Abate                \\[4pt]
        \textbf{Student ID:}     & 1716/16                    \\[4pt]
        \textbf{Program:}        & BSc Computer Science       \\[4pt]
        \textbf{Year / Semester:}& Year III, Semester II      \\[4pt]
        \textbf{Academic Year:}  & 2018/2026                  \\[4pt]
        \textbf{Submission Date:}& April 28, 2026             \\[4pt]
        \textbf{Advisor:}        & Mr.\ Birhane Bekele        \\
    \end{tabular}\\[1.5cm]
    \textit{Submitted in partial fulfillment of the requirements for\\
    Research Methods in Computer Science (CoSc3101)}
    \vfill
    {\small Department of Computer Science, Hawassa University Institute of Technology, Ethiopia}
\end{titlepage}

% ── Abstract ──────────────────────────────────────────────────────────────────
\begin{abstract}
\noindent
Deep learning-based Intrusion Detection Systems (DL-IDS) have become a cornerstone
of modern cybersecurity infrastructure; however, they remain critically vulnerable
to adversarial examples---carefully crafted perturbations that cause misclassification
while remaining imperceptible to human analysts. This proposal investigates the
robustness of DL-IDS models against white-box and black-box adversarial attacks and
proposes an \textit{Ensemble Adversarial Training (EAT)} framework that simultaneously
hardens multiple neural architectures. Using the CICIDS2017 and NSL-KDD benchmark
datasets, the study will quantify detection degradation under FGSM, PGD, and
Carlini--Wagner attacks, then evaluate whether the proposed EAT strategy restores
and exceeds baseline detection rates. Evaluation metrics include accuracy, F1-score,
False Negative Rate (FNR), and adversarial robustness score. This research directly
addresses a gap in the existing literature---the absence of a unified, dataset-agnostic
adversarial hardening approach for network intrusion detection---and contributes
a reproducible open-source framework for the cybersecurity community.
\end{abstract}

\tableofcontents
\newpage

% ══════════════════════════════════════════════════════════════════════════════
\section{Logical Foundation}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Title}

\textbf{Detecting and Mitigating Adversarial Attacks on Deep Learning-Based
Intrusion Detection Systems Using Ensemble Adversarial Training}

\subsection{Problem Statement}

\textbf{Ideal State.}
In an ideal cybersecurity environment, Deep Learning-based Intrusion Detection Systems
(DL-IDS) reliably detect malicious network traffic with high accuracy ($>$99\%),
remain robust against evolving evasion strategies, and maintain low false-negative
rates even when faced with deliberately obfuscated attack patterns.

\textbf{Reality.}
Current DL-IDS models---including convolutional neural networks (CNNs) and
recurrent neural networks (RNNs)---are susceptible to adversarial examples.
Studies show that FGSM-perturbed traffic samples can reduce detection accuracy
from over 98\% to below 45\% in standard benchmark environments \cite{yang2022}.
Despite the growing body of research on adversarial machine learning, existing
hardening methods are typically evaluated in isolation on a single architecture and
a single dataset, making their generalizability questionable for real-world
heterogeneous networks.

\textbf{Consequence.}
Without robust adversarial defenses, attackers can systematically exploit the
blind spots of deployed DL-IDS to infiltrate enterprise networks, exfiltrate
sensitive data, and disrupt critical infrastructure---with the global average
cost of a data breach reaching USD~4.45 million in 2023 \cite{ibm2023}.

\subsection{Research Objectives}

\textbf{General Objective.}
To design, implement, and evaluate an Ensemble Adversarial Training (EAT) framework
that improves the robustness of deep learning-based intrusion detection systems
against both white-box and black-box adversarial attacks.

\textbf{Specific SMART Objectives.}
\begin{enumerate}[label=\textbf{SO\arabic*.}, leftmargin=*, itemsep=4pt]
    \item \textbf{Quantify} the degradation in detection accuracy ($\Delta$Acc)
          of three DL-IDS architectures (CNN, LSTM, and Autoencoder) under FGSM,
          PGD, and Carlini--Wagner adversarial attacks on the CICIDS2017 dataset
          by \textit{Week~6} of the project timeline.
    \item \textbf{Implement} the proposed EAT framework by integrating adversarial
          samples from at least two attack methods into the training pipeline of
          each architecture and achieve a post-hardening F1-score $\geq 0.90$
          on the NSL-KDD test set by \textit{Week~10}.
    \item \textbf{Evaluate} and compare the EAT framework against three baseline
          defense strategies (standard adversarial training, feature squeezing, and
          input transformation) using accuracy, FNR, and adversarial robustness
          score, and produce a reproducible open-source implementation on GitHub
          by \textit{Week~14}.
\end{enumerate}

% ══════════════════════════════════════════════════════════════════════════════
\section{Literature Synthesis and State of the Art}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Synthesis Matrix}

Table~\ref{tab:synthesis} summarizes five peer-reviewed works from IEEE Xplore,
ACM Digital Library, and ScienceDirect (2022--2026).

\begin{table}[ht]
\centering
\caption{Synthesis Matrix of Related Work (2022--2026)}
\label{tab:synthesis}
\small
\renewcommand{\arraystretch}{1.4}
\begin{tabularx}{\textwidth}{>{\bfseries}p{2.2cm} p{2.5cm} p{2.2cm} p{2.2cm} p{3.2cm}}
\toprule
\textbf{Author(s)} & \textbf{Method} & \textbf{Dataset} & \textbf{Metric} & \textbf{Limitation} \\
\midrule
Yang et al.\ \cite{yang2022} &
FGSM + LSTM-based IDS &
CICIDS2017 &
Accuracy, FPR &
Only white-box attacks; single architecture evaluated \\

Han et al.\ \cite{han2023} &
GAN-based adversarial traffic generation &
NSL-KDD &
Detection rate, AUC &
Generative model prone to mode collapse; no defense proposed \\

Debicha et al.\ \cite{debicha2023} &
Adversarial training with PGD &
UNSW-NB15 &
F1-score, FNR &
Single-model defense; high computational cost; not tested on heterogeneous traffic \\

Li et al.\ \cite{li2024} &
Ensemble learning for IDS robustness &
CIC-DDoS2019 &
Precision, Recall &
Ensemble not hardened against adversarial inputs; no adversarial training applied \\

Sarhan et al.\ \cite{sarhan2022} &
Feature-based transfer learning IDS &
CICIDS2017, UNSW-NB15 &
Accuracy, F1 &
Transfer assumes similar feature distributions; adversarial robustness not evaluated \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Synthesis and Research Gap (approx.\ 300 words)}

The landscape of adversarial attacks on deep learning-based intrusion detection systems
has evolved rapidly since 2022. Yang et al.\ \cite{yang2022} demonstrated that even
a simple gradient-based perturbation (FGSM) can catastrophically reduce LSTM-IDS
accuracy on CICIDS2017, confirming that the threat is not merely theoretical.
Han et al.\ \cite{han2023} extended this line of inquiry by using Generative Adversarial
Networks (GANs) to synthesize realistic malicious traffic variants; however, their work
stops short of proposing a corresponding defense, leaving a critical gap between
attack modeling and practical mitigation.

On the defense side, Debicha et al.\ \cite{debicha2023} applied Projected Gradient
Descent (PGD) adversarial training to a single-model IDS on UNSW-NB15 and reported
meaningful improvements in F1-score. However, their approach is computationally
expensive and inherently architecture-specific---a single hardened model provides
no resilience if a different classifier is deployed in the same network. Li et al.\
\cite{li2024} recognized the performance benefits of ensemble learning for IDS
and demonstrated higher recall on DDoS datasets, yet their ensemble was not subjected
to any adversarial perturbations, leaving its robustness entirely unknown.

Sarhan et al.\ \cite{sarhan2022} proposed cross-dataset transfer learning for IDS
feature generalization, which is promising for deployment diversity but does not
address adversarial robustness at all.

The synthesis reveals a clear and consistent gap: \textit{no existing study evaluates
an adversarial hardening strategy that is simultaneously (a)~multi-architecture,
(b)~multi-dataset, and (c)~compared against more than one baseline defense}.
The proposed Ensemble Adversarial Training (EAT) framework directly addresses this
gap by jointly training CNN, LSTM, and Autoencoder models with adversarial samples
drawn from multiple attack methods and validating results on two independent benchmark
datasets. This positions the study as a meaningful contribution to adversarially
robust cybersecurity systems.

% ══════════════════════════════════════════════════════════════════════════════
\section{Technical Methodology and Experimental Design}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{System Architecture}

The proposed system follows a four-stage pipeline:

\begin{enumerate}[label=\textbf{Stage~\arabic*:}, leftmargin=*, itemsep=4pt]
    \item \textbf{Data Ingestion and Preprocessing.}
          Raw network flow records from CICIDS2017 and NSL-KDD are ingested,
          cleaned (null removal, label encoding), and normalized using
          Min-Max scaling. Features are selected via mutual information ranking
          to retain the top-40 most discriminative attributes.

    \item \textbf{Adversarial Attack Generation.}
          Three attack algorithms---FGSM, PGD, and Carlini--Wagner (C\&W)---are
          applied to the preprocessed test set using the Adversarial Robustness
          Toolbox (ART v1.16). Both white-box (full model access) and black-box
          (transfer-based) attack scenarios are simulated.

    \item \textbf{Ensemble Adversarial Training (EAT).}
          Three base classifiers (CNN, LSTM, Autoencoder) are trained on an
          augmented dataset comprising original samples plus adversarial samples
          from FGSM and PGD. A soft-voting ensemble aggregates their output
          probabilities. The training loop iterates adversarial sample generation
          and retraining for up to five epochs of augmentation.

    \item \textbf{Evaluation and Comparison.}
          The hardened ensemble is benchmarked against three baselines
          (standard adversarial training, feature squeezing, input transformation)
          using a held-out test set. Results are reported per attack type and
          per dataset.
\end{enumerate}

\subsection{Tool Justification}

\begin{itemize}[itemsep=4pt]
    \item \textbf{Python 3.11 + PyTorch 2.2} --- Chosen over TensorFlow because
          PyTorch's dynamic computation graph simplifies custom adversarial training
          loops and is the dominant framework in adversarial ML research (used by
          ART, CleverHans, and Foolbox).
    \item \textbf{IBM Adversarial Robustness Toolbox (ART v1.16)} --- Framework-agnostic
          library supporting FGSM, PGD, and C\&W out of the box with a consistent API;
          preferred over CleverHans (less actively maintained) and Foolbox (fewer
          white-box attack implementations).
    \item \textbf{Scikit-learn 1.4} --- Used for preprocessing pipelines, feature
          selection (mutual information), and baseline classifier comparison.
    \item \textbf{Weights \& Biases (WandB)} --- Experiment tracking and hyperparameter
          logging, enabling fully reproducible runs linked to the GitHub repository.
    \item \textbf{Jupyter Notebook} --- Provides a literate programming environment
          for incremental development; final scripts are exported to \texttt{.py}
          files for reproducibility.
\end{itemize}

\subsection{Data Strategy}

\textbf{Datasets:}
\begin{itemize}[itemsep=2pt]
    \item \textbf{CICIDS2017} (Canadian Institute for Cybersecurity) --- 2.8~million
          records, 78 features, covering seven attack categories (DoS, DDoS, Brute
          Force, Infiltration, etc.). Available on IEEE DataPort and the CIC website.
    \item \textbf{NSL-KDD} (UCI Machine Learning Repository) --- 125,973 training
          records, 41 features, four attack classes. A cleaned successor to KDD Cup 99
          with no duplicate records.
\end{itemize}

\textbf{Preprocessing Steps:}
\begin{enumerate}[itemsep=2pt]
    \item Remove records with infinite or NaN values.
    \item Encode categorical labels (attack type) as integers.
    \item Apply Min-Max normalization to all numerical features ($[0,1]$ range).
    \item Select top-40 features using mutual information with respect to the label.
    \item Split into 70\% training, 15\% validation, 15\% test (stratified by class).
    \item Generate adversarial augmentations only on the training split to prevent
          data leakage.
\end{enumerate}

\subsection{Evaluation Metrics}

The following metrics will be used to evaluate model performance:

\begin{align}
    \text{Accuracy} &= \frac{TP + TN}{TP + TN + FP + FN} \\[6pt]
    \text{F1-score} &= \frac{2 \cdot \text{Precision} \cdot \text{Recall}}
                           {\text{Precision} + \text{Recall}} \\[6pt]
    \text{False Negative Rate (FNR)} &= \frac{FN}{FN + TP} \\[6pt]
    \text{Adversarial Robustness Score (ARS)} &= 1 - \frac{\text{Acc}_{\text{clean}} -
                                               \text{Acc}_{\text{adv}}}
                                               {\text{Acc}_{\text{clean}}}
\end{align}

Success is defined as: post-hardening F1-score $\geq 0.90$, FNR $\leq 0.08$,
and ARS $\geq 0.85$ on both benchmark datasets.

% ══════════════════════════════════════════════════════════════════════════════
\section{Project Management and Ethics}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{Timeline --- Gantt Chart}

Figure~\ref{fig:gantt} shows the project timeline across 16 weeks (four phases).

\begin{figure}[ht]
\centering
\caption{Project Gantt Chart}
\label{fig:gantt}
\begin{ganttchart}[
    hgrid,
    vgrid={*{6}{draw=none}, dotted},
    x unit=0.62cm,
    y unit title=0.55cm,
    y unit chart=0.52cm,
    title label font=\scriptsize,
    bar label font=\scriptsize,
    group label font=\small\bfseries,
    bar/.style={fill=blue!50, rounded corners=2pt},
    group/.style={fill=blue!80!black},
    milestone/.style={fill=red, rounded corners=2pt},
    bar height=0.5,
    title height=1
]{1}{16}
    \gantttitle{Project Timeline (Weeks)}{16} \\
    \gantttitlelist{1,...,16}{1} \\

    \ganttgroup{Phase 1: Literature Review}{1}{4} \\
    \ganttbar{Literature search \& reading}{1}{3} \\
    \ganttbar{Synthesis matrix construction}{3}{4} \\

    \ganttgroup{Phase 2: Data \& Setup}{4}{7} \\
    \ganttbar{Dataset acquisition \& cleaning}{4}{5} \\
    \ganttbar{Environment setup (PyTorch, ART)}{5}{6} \\
    \ganttbar{Baseline model implementation}{6}{7} \\

    \ganttgroup{Phase 3: Experiments}{7}{12} \\
    \ganttbar{Adversarial attack generation}{7}{9} \\
    \ganttbar{EAT framework implementation}{9}{11} \\
    \ganttbar{Baseline defense comparison}{11}{12} \\

    \ganttgroup{Phase 4: Evaluation \& Writing}{12}{16} \\
    \ganttbar{Results analysis \& visualization}{12}{13} \\
    \ganttbar{Final report \& GitHub cleanup}{13}{15} \\
    \ganttbar{Submission preparation}{15}{16} \\
    \ganttmilestone{Final Submission}{16}
\end{ganttchart}
\end{figure}

\subsection{Ethical Considerations}

\textbf{Data Privacy (GDPR Compliance).}
Both CICIDS2017 and NSL-KDD are publicly available benchmark datasets derived
from simulated network environments. They contain no real user identities, personal
communications, or private IP addresses traceable to individuals. As such, the
research does not involve the collection of personal data and is not subject to
GDPR consent requirements. However, any future deployment of the proposed IDS
in a live network must comply with GDPR Article~6 (lawfulness of processing) and
Article~32 (security of processing), particularly regarding the logging of user
traffic data.

\textbf{Algorithmic Bias.}
The training datasets were collected from controlled laboratory environments and
may not represent the full diversity of real-world network traffic (e.g., traffic
from IoT-dense African networks may differ significantly from North American campus
traffic). Models trained solely on CICIDS2017 may exhibit lower recall on traffic
patterns not represented in that dataset. To mitigate this, the study uses two
independent datasets and reports per-class F1-scores to surface any class-level
bias.

\textbf{Environmental Impact.}
Deep learning training is energy-intensive. All experiments will be run on a single
GPU (NVIDIA RTX 3060, 12~GB VRAM) with early stopping enabled to minimize unnecessary
computation. Estimated energy consumption will be reported in the final paper using
the \texttt{codecarbon} library, promoting transparency around the environmental
cost of the research.

% ══════════════════════════════════════════════════════════════════════════════
\section{Version Control and Documentation}
% ══════════════════════════════════════════════════════════════════════════════

\subsection{GitHub Repository}

The project repository is hosted publicly at:\\
\url{https://github.com/Dagi7d/ARP-Poisoning-Detection-GBM}

\textbf{Repository Structure:}
\begin{verbatim}
ARP-Poisoning-Detection-GBM/
├── README.md
├── main.tex
└── references.bib
\end{verbatim}

\subsection{README.md Contents}

The \texttt{README.md} will include:
\begin{enumerate}[itemsep=2pt]
    \item An abstract of the research proposal (mirroring the abstract above).
    \item Step-by-step installation instructions for all dependencies.
    \item Instructions to download and place datasets in \texttt{data/raw/}.
    \item Commands to reproduce all experiments.
\end{enumerate}

Sample installation block (to appear in README):
\begin{verbatim}
# Clone the repository
git clone https://github.com/Dagi7d/ARP-Poisoning-Detection-GBM.git
cd adversarial-ids-eat

# Create virtual environment
python -m venv venv
source venv/bin/activate      # Windows: venv\Scripts\activate

# Install dependencies
pip install -r requirements.txt

# Run baseline evaluation
python src/models/cnn_ids.py --dataset cicids2017

# Run EAT framework
python src/eat_framework.py --attack fgsm pgd --epochs 5
\end{verbatim}

\subsection{BibTeX}

All references are managed in \texttt{references.bib} using BibTeX. No manual
bibliography entries are used. Citations in this document are generated with
the \texttt{cite} package and the \texttt{IEEEtran} bibliography style.

% ── References ────────────────────────────────────────────────────────────────
\newpage
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
